Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-06-04 submission/manuscript.Rmd
smooth out the intro

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. The envrionment has been attributed as an important risk factor for CRC [@Lichtenstein2000]. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011]. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment [@test_OKeefe2016]. In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects. Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC there is a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, by increasing the adenoma tumor cohort it we will be able to identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases. It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors. This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group will be at a cross sectional point in time while the second group will be before (pre-) and after (post-)treatment. Additionally, we also will assess whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma, use PICRUSt and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyze how well 16S rRNA gene sequencing can predict SCFA concentrations. This extensive investigation will provide additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against tumorigenesis.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the envrionment has been attributed as an important and large risk factor for CRC [@Lichtenstein2000]. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colons to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @Kostic2011]. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a potential treatment option.

Specific SCFAs, like butyrate, have shown positive results for CRC treatment within model systems [@test_OKeefe2016]. Butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects. For example, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals with normal colons [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, by increasing the adenoma tumor cohort it we will identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases. It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors (adenoma or carcinoma). This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we also assessed whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], used PICRUSt [@Langille2013] and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyzed how well 16S rRNA gene sequencing can predict SCFA concentrations. This  investigation provides additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against colon tumorigenesis.
